---
title: "R for Data Science: Data Visualisation"
author: "tony fu"
date: "8/30/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


### Installation
mpg contains observations collected by the US Environmental Protection Agency on 38 models of car.
 `displ`, a car’s engine size, in litres.
 `hwy`, a car’s fuel efficiency on the highway, in miles per gallon (mpg). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance.
```{r}
#install.packages("tidyverse")
library(tidyverse)

```{r}
mpg
```

#### 3.2.2 Creating a ggplot
The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy).
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


#### 3.3 Aesthetic mappings
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy,color = class))
  
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
#> Warning: Using size for a discrete variable is not advised.
```



```{r}
ggplot(data = mpg) +
   geom_point(mapping = aes(x = displ, y = hwy), color = "red")
```
#### 3.32 Aesthetic mappings
### how 
#### 3.5 Facets
One way to add additional variables is with aesthetics. Another way, particularly useful for categorical variables, is to split your plot into facets, subplots that each display one subset of the data.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
#### 3.6Geometric objects
A geom is the geometrical object that a plot uses to represent data. People often describe plots by the type of geom that the plot uses. For example, bar charts use bar geoms, line charts use line geoms, boxplots use boxplot geoms, and so on.

```{r}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

```

Display multiple geoms in the same plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
This, however, introduces some duplication in our code. Imagine if you wanted to change the y-axis to display cty instead of hwy. You’d need to change the variable in two places, and you might forget to update one. You can avoid this type of repetition by passing a set of mappings to ggplot(). ggplot2 will treat these mappings as global mappings that apply to each geom in the graph. In other words, this code will produce the same plot as the previous code:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

### 3.7 Statistical transformations
Bar charts seem simple, but they are interesting because they reveal something subtle about plots. \n
The following chart displays the total number of diamonds in the diamonds dataset, grouped by cut. The diamonds dataset comes in ggplot2 and contains information about ~54,000 diamonds, including the price, carat, color, clarity, and cut of each diamond. The chart shows that more diamonds are available with high quality cuts than with low quality cuts.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

#### 3.9 Coordinate systems
`coord_flip()` switches the x and y axes. This is useful (for example), if you want horizontal boxplots. It’s also useful for long labels: it’s hard to get them to fit without overlapping on the x-axis.
```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```
`coord_quickmap()` sets the aspect ratio correctly for maps. This is very important if you’re plotting spatial data with ggplot2 (which unfortunately we don’t have the space to cover in this book).

```{r}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "blue")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```



