---
title: 'R for DataScience:  Data transformation'
author: "tony fu"
date: "9/1/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Data transformation

Install the package:
```{r}
#install.packages("nycflights13")
```
Load the required packages
```{r}
library(nycflights13)
library(tidyverse)
```

```{r}
flights
```
Use with function `filter()` ,allows you to subset observations based on their values. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. For example, we can select all flights on January 1st with:
```{r}
filter(flights, month == 1, day == 1)

```
dplyr functions never modify their inputs, so if you want to save the result, you’ll need to use the assignment operator, `<-`:
```{r}
jan1 <- filter(flights, month == 1, day == 1)
```

There’s another common problem you might encounter when using ==: floating point numbers. These results might surprise you!
```{r}
sqrt(2) ^ 2 == 2 
1 / 49 * 49 == 1
```
Computers use **finite precision** arithmetic (they obviously can’t store an infinite number of digits!) so remember that every number you see is an approximation. Instead of relying on ==, use near():
```{r}
near(sqrt(2) ^ 2,  2)
#> [1] TRUE
near(1 / 49 * 49, 1)
#> [1] TRUE
```

For`NA`, a special case is that `NA == NA` : 
```{r}
NA == NA
#> [1] NA
```
It’s easiest to understand why this is true with a bit more context:
```{r}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
#> [1] NA
# We don't know!
```
If you want to determine if a value is missing, use `is.na()`:
```{r}
is.na(10)
is.na(x)
is.na(x==y)
```

`filter()` only includes rows where the condition is **TRUE**; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:
```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
#> # A tibble: 1 x 1
#>       x
#>   <dbl>
#> 1     3
filter(df, is.na(x) | x > 1)
#> # A tibble: 2 x 1
#>       x
#>   <dbl>
#> 1    NA
#> 2     3
```

#### Arrange rows with `arrange()`
`arrange()` works similarly to `filter()` except that instead of selecting rows, it changes their order. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:
```{r}
arrange(flights, year, month, day)
#> # A tibble: 336,776 x 19
#>    year month   day dep_time sched_dep_time dep_delay arr_time sched_arr_time
#>   <int> <int> <int>    <int>          <int>     <dbl>    <int>          <int>
#> 1  2013     1     1      517            515         2      830            819
#> 2  2013     1     1      533            529         4      850            830
#> 3  2013     1     1      542            540         2      923            850
#> 4  2013     1     1      544            545        -1     1004           1022
#> 5  2013     1     1      554            600        -6      812            837
#> 6  2013     1     1      554            558        -4      740            728
#> # … with 336,770 more rows, and 11 more variables: arr_delay <dbl>,
#> #   carrier <chr>, flight <int>, tailnum <chr>, origin <chr>, dest <chr>,
#> #   air_time <dbl>, distance <dbl>, hour <dbl>, minute <dbl>, time_hour <dttm>
```
Use `desc()` to re-order by a column in descending order:
```{r}
arrange(flights, desc(dep_delay))
```
Note that missing values are always sorted at the end:
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
#> # A tibble: 3 x 1
#>       x
#>   <dbl>
#> 1     2
#> 2     5
#> 3    NA
arrange(df, desc(x))
#> # A tibble: 3 x 1
#>       x
#>   <dbl>
#> 1     5
#> 2     2
#> 3    NA
```

#### Exercises
1.How could you use `arrange()` to sort all missing values to the start? (Hint: use is.na()).
My ans: 
```{r}
arrange(df,desc(is.na(df)))
```
Solotions:
To put NA values first, we can add an indicator of whether the column has a missing value. Then we sort by the missing indicator column and the column of interest. For example, to sort the data frame by departure time (dep_time) in ascending order but NA values first, run the following.
```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```

2.Sort flights to find the most delayed flights. Find the flights that left earliest.
my ans:
the most delayed flight
```{r}
arrange(flights, desc(dep_delay))
```
the flights that left earliest:
```{r}
arrange(flights,dep_delay)
```



#### Select columns with `select()`
It’s not uncommon to get datasets with hundreds or even thousands of variables. In this case, the first challenge is often narrowing in on the variables you’re actually interested in. `select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.
```{r}
# Select columns by name
select(flights, year, month, day)
select(flights, year:day)
select(flights, -(year:day))

```
Another option is to use select() in conjunction with the everything() helper. This is useful if you have a handful of variables you’d like to move to the start of the data frame.
```{r}
select(flights, time_hour, air_time, everything())

```

#### Exercices 
1.What happens if you include the name of a variable multiple times in a `select()` call?
The `select()` call **ignores** the duplication. Any duplicated variables are only included once, in the first location they appear. The `select()` function does not raise an error or warning or print any message if there are duplicated variables.
```{r}
select(flights, year, month, day, year, year)
```
This behavior is useful because it means that we can use `select()` with `everything()` in order to easily **change the order** of columns without having to specify the names of all the columns.
```{r}
select(flights, arr_delay, everything())
```

2.What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?
Ans: The `one_of()` function selects variables with a character vector rather than unquoted variable name arguments. This function is useful because it is easier to programmatically generate character vectors with variable names than to generate unquoted variable names, which are easier to type.

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```
**Note**: In the most recent versions of dplyr, `one_of` has been deprecated in favor of two functions: `all_of()` and `any_of()`. These functions behave similarly if all variables are present in the data frame.

```{r}
select(flights, any_of(vars))
select(flights, all_of(vars))
```

These functions **differ** in their **strictness**. The function `all_of()` will raise an error if one of the variable names is not present, while `any_of()` will ignore it.
```
vars2 <- c("year", "month", "day", "variable_not_in_the_dataframe")
select(flights, all_of(vars2))
#> Error: Can't subset columns that don't exist.
#> ✖ Column `variable_not_in_the_dataframe` doesn't exist.
```

```
select(flights, any_of(vars2))
#> # A tibble: 336,776 x 3
#>    year month   day
#>   <int> <int> <int>
#> 1  2013     1     1
#> 2  2013     1     1
#> 3  2013     1     1
#> 4  2013     1     1
#> 5  2013     1     1
#> 6  2013     1     1
#> # … with 336,770 more rows
```
####  Add new variables with `mutate()`
Besides selecting sets of existing columns, it’s often useful to add new columns that are functions of existing columns. That’s the job of `mutate()`. in the below example, we could create `gain` and `speed` on the existing dataframe.
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```
Note that you can refer to columns that you’ve just created:
```
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
#> # A tibble: 336,776 x 10
#>    year month   day dep_delay arr_delay distance air_time  gain hours
#>   <int> <int> <int>     <dbl>     <dbl>    <dbl>    <dbl> <dbl> <dbl>
#> 1  2013     1     1         2        11     1400      227    -9  3.78
#> 2  2013     1     1         4        20     1416      227   -16  3.78
#> 3  2013     1     1         2        33     1089      160   -31  2.67
#> 4  2013     1     1        -1       -18     1576      183    17  3.05
#> 5  2013     1     1        -6       -25      762      116    19  1.93
#> 6  2013     1     1        -4        12      719      150   -16  2.5 
#> # … with 336,770 more rows, and 1 more variable: gain_per_hour <dbl>
```
If you only want to keep the new variables, use `transmute()`, from here we could see that the original columns are no longer in this new table 
```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

#### Excercises
1.Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
My Ans:
since `dept_time` means : Actual departure and arrival times (format `HHMM` or `HMM`), local tz.
```{r}
depart_info <- select(flights,dep_time,sched_dep_time)
mutate(depart_info,
      dep_time_min = ((dep_time %/%100)*60 + dep_time %% 100) %% 1440,
      sched_dep_time_min =  ((sched_dep_time %/%100)*60 + sched_dep_time %% 100) %% 1440
      
)
```
Better Solutions:
```{r}
time2mins <- function(x) {
  (x %/% 100 * 60 + x %% 100) %% 1440
}

flights_times <- mutate(flights,
  dep_time_mins = time2mins(dep_time),
  sched_dep_time_mins = time2mins(sched_dep_time)
)
# show only the relevant columns
select(
  flights_times, dep_time, dep_time_mins, sched_dep_time,
  sched_dep_time_mins
)

```

Compare air_time with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
select(flights,arr_time)
```




